Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-01-17 E-Hypergraphs-LICS24.tex
S6 update

We use the notation $n \setminus m$ to denote the discrete hypergraph with $n - m$ nodes,  in particular with removal of the nodes of $m$ removed from $n$ when nodes of $m$ are identified within $n$.
\begin{definition}[Category of e-hypergraphs with interfaces]
    The category of \textit{e-hypergraphs with interfaces} $\Ecospans$ has discrete e-hypergraphs as objects,  with homsets $\Ecospans(n,m)$ consisting of isomorphism classes of \textit{extended cospans}, defined as follows:  
    \[
    n \xrightarrow{f_{ext}} n' \xrightarrow{f_{int}} \mathcal{G} \xleftarrow{g_{int}} m' \xleftarrow{g_{ext}} m
    \]
    where $\mathcal{G}$ is an e-hypergraph,  and $n, n', m, m'$ are discrete e-hypergraphs,  $f_{ext},f_{int},g_{ext},g_{int}$ are monomorphisms in $\catname{EHyp(\Sigma)}$,  and the image of $f_{ext};f_{int}$ and of $g_{ext};g_{int}$ consist exclusively of top-level vertices,  and that that vertices in the strictly internal interface (defined below) are not top-level.

This justifies calling $n$  \textit{external input interfaces} and $n'$ \textit{internal input (output) interfaces}.  We call $n' \setminus f_{ext}(n)$  the \textit{strictly internal input interfaces}.  We do analogously for the \emph{output interfaces},  with respect to $m$,  $m'$ and $m' \setminus g_{ext}(m)$.  
We occasionally conflate $f_{ext}$ with $f_{ext};f_{int}$ when it is clear from context,  and also conflate $n, n', m$ and $m'$ with their images in $\mathcal{G}$. 
Given an edge $e \in E_\mathcal{G}$ such that $l(e) = \bot$,  we call the \textit{inputs of $e$} the intersection of the strictly internal input interface of $\mathcal{G}$ with the immediate successors of $e$,  and analogously for the \emph{outputs of $e$}.

@@

CHRIS: INTRO.

@@

We wish to specify a set of DPOI rewrite rules corresponding to those in Figure \ref{fig:string-equations}.  In order to do this,  we specify a function which maps  uninterpreted $\Sigma^+$-terms into $\WellTypedMdaEcospans$ morphisms,  and which also extends the expected interpretation of terms in $\WellTypedMdaEcospans$ (as given by the free functor from $\textbf{PROP}(\Sigma)$ induced by the interpretation of $\Sigma$ as itself,  detailed in \cite{bonchi_string_2022-2}). 
\begin{definition}[Interpretation function] 
%Let $\llbracket - \rrbracket: \textbf{PROP}(\Sigma) \to \WellTypedMdaEcospans$ be the unique SMC-functor induced by the interpretation sending $\Sigma$ to itself.  
Let $\llbracket - \rrbracket_{n,m}: \Sigma^+(n,m) \to \WellTypedMdaEcospans(n,m)$ be the function defined by induction on the type derivation of $\Sigma$-terms by 
\begin{align*}
	\llbracket \id \rrbracket_{n,m} &= \id  &	\llbracket \phi ; \psi \rrbracket_{n,m} &= \llbracket \phi \rrbracket_{n,m} ; \llbracket \psi \rrbracket_{n,m} \\
	\llbracket 1 \rrbracket_{n,m} &= 1 & \llbracket f \otimes g \rrbracket_{n,m} &= \llbracket f \rrbracket^+_{n,m} \otimes \llbracket g \rrbracket_{n,m} \\
	\llbracket \sym \rrbracket_{n,m} &= \sym  & \llbracket c \rrbracket_{n,m} &= c   \\ 
	\llbracket f+g \rrbracket_{n,m} &= \llbracket f \rrbracket_{n,m} + \llbracket g \rrbracket^+_{n,m}
\end{align*}
where $c \in \Sigma$ and $+$ in $\WellTypedMdaEcospans$ is defined by Figure \ref{fig:A+B}.  We will omit the subscripts in $\llbracket - \rrbracket_{n,m}$ where it is clear from context. 
\end{definition}

Let the function $\llbracket - \rrbracket$ extend in the obvious way to apply to equations (\textit{i.e.}, pairs of $\Sigma^+$-terms). 
% \update{below definition depends on whether we require the homomorphism to reflect the conflicts}
CHRIS: COMMENT THAT WE QUOTIENT TO PHRASE CATEGORICALLY OUR RESULTS. 
\begin{definition}[Quotient by rewrites]  
Given a set of equations $\mathcal{E}$,  we denote by $\WellTypedMdaEcospans/\mathcal{E}$ the category $\WellTypedMdaEcospans$ quotiented by the following relation. 
\[
	f \sim g \quad \text{if} \quad f \Rrightarrow^{+,*}_{\llbracket \mathcal E \rrbracket} g ~ . 
\]
\end{definition}
In particular,  let $\mathcal{S}$ be the set equations of Figure \ref{fig:string-equations},  \textit{i.e.},  the semilattice equations over $\Sigma^+$-terms. 
This quotient in fact turns the interpretation function into a proper functor $\llbracket - \rrbracket: \textbf{PROP}^+(\Sigma) \to \WellTypedMdaEcospans/{\mathcal{S}}$,  which can further be quotiented to account for extra equations $\mathcal E$ arising in symmetric monoidal theories,  as below. 
\[
	\llbracket - \rrbracket_{\mathcal E}: \textbf{SMT}^+(\Sigma, \mathcal E) \to \WellTypedMdaEcospans/{\mathcal{S,E}}
\]

Note that not all pushouts (or pushout complements) in $\textbf{MEHypI}(\Sigma)$ are guaranteed to exist: however,  pushouts (and pushout complements) for rewrites generated from $\Sigma$-equations and those generated from the semilattice equations on $\Sigma^+$-terms do exist.  
%Let $\mathcal E$ be an arbitrary set of $\Sigma$-equations. 
Thus we have the following result. 
\begin{proposition}[Soundness]
\label{prop:soundness}
The category $\WellTypedMdaEcospans/{\mathcal{S}}$ is a semilattice-enriched PROP. 
\end{proposition}
It immediately follows that $\WellTypedMdaEcospans/{\mathcal{S,E}}$ is a semilattice-enriched PROP,  and $\llbracket - \rrbracket_{\mathcal E}$ is the free semilattice-enriched functor from $\textbf{SMT}^+(\Sigma, \mathcal E)$ induced by the interpretation of $\Sigma$ as itself. 
Note that the symmetric monoidal equations are absorbed in the hypergraph representation,  but the semilattice equations are covered by the fact we have quotiented our representation by certain rewrites.

The equations of Figure~\ref{fig:string-equations} -- excepting the commutativity equation -- can all be directed in an obvious way to form a terminating rewrite system,  which results in a cospan in \textit{weak normal form}. 
\begin{proposition}[Weak normal form]
\label{prop:wnormal_form}
For each cospan $f$ in ${\WellTypedMdaEcospans}/{\mathcal{S}}$ there is a \textit{weak normal form} such that 
\[
	f = f_1 + \ldots + f_n
\]
 such that each $f_i$ contains no hierarchical edges,  and for all $i \neq j$ we have $f_i \neq f_j$.
\end{proposition}    
Note that at similar argument can be used to rewrite $\Sigma^+$-terms into a similar form by using only the semilattice-enrichment equations.

This normal form can be used to prove the main theorem of the paper: the (full) completeness of our combinatorial representation of $\textbf{SMT}^{+}(\Sigma, \mathcal {E} )$. 
\begin{theorem}[Completeness]
We have the following equivalence of categories
\[
	\textbf{SMT}^{+}(\Sigma, \mathcal {E} ) \cong \WellTypedMdaEcospans / \mathcal{S,E}~.
\]
\end{theorem}
\begin{proof}
The equivalence is given by the functor $\llbracket - \rrbracket$.  First, note that the functor is surjective on objects,  so it suffices to prove the functor is both full and faithful.  In each case,  the result follows by a normal form argument in combination with relying on the equivalence $(*)$ $\textbf{SMT}(\Sigma, \mathcal E) \cong \textbf{MHypI}(\Sigma)/\mathcal{E}$ of symmetric monoidal theories and (appropriately quotiented) plain MDA cospans which is given by the restriction of $\llbracket - \rrbracket$ to $\textbf{SMT}(\Sigma, \mathcal E)$ as a subcategory of $\textbf{SMT}^{+}(\Sigma, \mathcal {E} )$ (note that cospans embed in extended cospans when the internal interface is isomorphic to the external interface); see Theorem 35 of ~\cite{bonchi_string_2022-2}.

@@

Unfolding the definitions,  we can spell out the previous theorem in terms of DPO-rewriting. 
\begin{corollary}[Completeness of rewriting]
For any $f, g \in \textbf{SMT}^{+}(\Sigma, \mathcal{E})$,  we have
\[
	f = g \quad \iff \quad \llbracket f \rrbracket \Rrightarrow^{+,*}_{\mathcal{S}, \mathcal{E}} \llbracket g \rrbracket~ . 
\]
 \end{corollary}
%
%\begin{theorem}[Theorem 35~\cite{bonchi_string_2022-2}]
%
%    $a \Rightarrow_{R} b$ iff $\llangle {}^{\ulcorner} a {}^{\urcorner} \rrangle \Rrightarrow_{\llangle {}^\ulcorner R {}^\urcorner \rrangle} \llangle {}^{\ulcorner}b {}^{\urcorner} \rrangle$.
%    Where $\llangle \cdot \rrangle : \textsf{PROP}(\Sigma) + \textsf{Frob} \to \catname{Csp_{D}(Hyp_{\Sigma})}$ and $\Rrightarrow$ is a convex rewriting relation for morphisms in $\catname{Hyp_{\Sigma}}$.
%\end{theorem}
%
%\begin{proposition}
%    $\llangle {}^{\ulcorner} a {}^{\urcorner} \rrangle \Rrightarrow_{\llangle {}^\ulcorner R {}^\urcorner \rrangle} \llangle {}^{\ulcorner}b {}^{\urcorner} \rrangle$ iff $[\mathcal{J}(a)] \Rrightarrow_{[R]}^{+} [\mathcal{J}(b)]$.
%     Note that $a$ and $b$ are from $\textsf{PROP}(\Sigma)$.
%\end{proposition}
%
%\begin{theorem}
%    $a \Rightarrow_{\langle l, r \rangle}^{+} b$ in $\textsf{PROP}^{+}$ iff $[a] \Rrightarrow_{\langle [l],[r] \rangle} [b]$ in $\sfrac{\WellTypedMdaEcospans}{\mathcal{S}}$.
%\end{theorem}
CHRIS: CONCLUDING STATEMENT.  COMMENT ABOUT GENERALIZATION TO DIRECTED REWRITES.

>>>>

We use the notation $n \setminus m$ to denote the discrete e-hypergraph with $n - m$ nodes,  in particular with the nodes of $m$ removed from $n$ when nodes of $m$ is a sub-e-hypergraph of $n$.
\begin{definition}[Category of e-hypergraphs with interfaces]
    The category of \textit{e-hypergraphs with interfaces} $\Ecospans$ has discrete e-hypergraphs as objects,  with homsets $\Ecospans(n,m)$ consisting of isomorphism classes of \textit{extended cospans}, defined as follows:  
    \[
    n \xrightarrow{f_{ext}} n' \xrightarrow{f_{int}} \mathcal{G} \xleftarrow{g_{int}} m' \xleftarrow{g_{ext}} m
    \]
    where $\mathcal{G}$ is an e-hypergraph,  and $n, n', m, m'$ are discrete e-hypergraphs,  $f_{ext},f_{int},g_{ext},g_{int}$ are monomorphisms in $\catname{EHyp(\Sigma)}$,  and the image of $f_{ext};f_{int}$ and of $g_{ext};g_{int}$ consist exclusively of top-level vertices,  and that that vertices in the strictly internal interface (defined below) are not top-level.

This justifies calling $n$  \textit{external input interfaces} and $n'$ \textit{internal input (output) interfaces}.  We call $n' \setminus f_{ext}(n)$  the \textit{strictly internal input interfaces}.  We do analogously for the \emph{output interfaces},  with respect to $m$,  $m'$ and $m' \setminus g_{ext}(m)$.  
We occasionally conflate $f_{ext}$ with $f_{ext};f_{int}$ when it is clear from context,  and also conflate $n$ and $m$ with their images in $n'$ and $m'$,  and likewise $m, m, n', m'$ with their images in $\mathcal{G}$. 
Given an edge $e \in E_\mathcal{G}$ such that $l(e) = \bot$,  we call the \textit{inputs of $e$} the intersection of the strictly internal input interface of $\mathcal{G}$ with the immediate successors of $e$,  and analogously for the \emph{outputs of $e$}.

@@

This section contains the main technical results of the paper,  building on those of \cite{bonchi_string_2022-2}. 
We first construct an interpretation of terms generating $\textbf{PROP}^+(\Sigma)$ in $\WellTypedMdaEcospans$.  
As expected,  the SMC equations are factored out in our representation of terms in $\WellTypedMdaEcospans$,  but the semilattice equations are not.  
Instead,  we implement the semilattice equations via EDPOI rewrites,  leading to the following soundness and completeness result: for any morphisms $f$ and $g$ in $\textbf{SMT}(\Sigma, \mathcal E)$,  $f = g$  if and only if there exists a sequence of EDPOI-rewrites --- each induced by either a structural semilattice equation or $\mathcal E$ --- between their combinatorial representations.

@@

We wish to specify a set of EDPOI rewrite rules corresponding to those in Figure \ref{fig:string-equations}.  In order to do this,  we specify a function which maps  uninterpreted $\Sigma^+$-terms into $\WellTypedMdaEcospans$ morphisms,  and which also extends the expected interpretation of $\Sigma$-terms in $\WellTypedMdaEcospans$ (as given by the free functor from $\textbf{PROP}(\Sigma)$ induced by the interpretation of $\Sigma$ as itself,  detailed in \cite{bonchi_string_2022-2}). 
\begin{definition}[Interpretation function] 
%Let $\llbracket - \rrbracket: \textbf{PROP}(\Sigma) \to \WellTypedMdaEcospans$ be the unique SMC-functor induced by the interpretation sending $\Sigma$ to itself.  
Let $\llbracket - \rrbracket_{n,m}: \Sigma^+(n,m) \to \WellTypedMdaEcospans(n,m)$ be the function defined by induction on the type derivation of $\Sigma$-terms by 
\begin{align*}
	\llbracket \id \rrbracket_{n,m} &= \id  &	\llbracket \phi ; \psi \rrbracket_{n,m} &= \llbracket \phi \rrbracket_{n,m} ; \llbracket \psi \rrbracket_{n,m} \\
	\llbracket 1 \rrbracket_{n,m} &= 1 & \llbracket f \otimes g \rrbracket_{n,m} &= \llbracket f \rrbracket^+_{n,m} \otimes \llbracket g \rrbracket_{n,m} \\
	\llbracket \sym \rrbracket_{n,m} &= \sym  & \llbracket c \rrbracket_{n,m} &= c   \\ 
	\llbracket f+g \rrbracket_{n,m} &= \llbracket f \rrbracket_{n,m} + \llbracket g \rrbracket^+_{n,m}
\end{align*}
where $c \in \Sigma$ and $+$ in $\WellTypedMdaEcospans$ is defined by Figure \ref{fig:A+B}.  We will omit the subscripts in $\llbracket - \rrbracket_{n,m}$ where it is clear from context. 
\end{definition}

In order to phrase our soundness and completeness results categorically,  we define the following quotient of $\WellTypedMdaEcospans$ by EDPOI rewrites.  This means that soundness is expressed as the extension of $\llbracket - \rrbracket$ to a semilattice-enriched SMC functor from $\textbf{SMT}^+(\Sigma, \mathcal E)$,  and completeness is expressed as the faithfulness of this functor.  In fact,  we prove that the functor is also full,  giving an equivalence of categories.

We define the appropriate notion of quotient by EDPOI rewrites below.  Let the function $\llbracket - \rrbracket$ extend in the obvious way to apply to equations (\textit{i.e.}, pairs of $\Sigma^+$-terms). 
% \update{below definition depends on whether we require the homomorphism to reflect the conflicts}
\begin{definition}[Quotient by rewrites]  
Given a set of equations $\mathcal{E}$,  we denote by $\WellTypedMdaEcospans/\mathcal{E}$ the category $\WellTypedMdaEcospans$ quotiented by the following relation. 
\[
	f \sim g \quad \text{if} \quad f \Rrightarrow^{*}_{\llbracket \mathcal E \rrbracket} g ~ . 
\]
\end{definition}

Note that not all pushouts (or pushout complements) in $\textbf{MEHypI}(\Sigma)$ are guaranteed to exist: however,  pushouts (and pushout complements) for rewrites generated from $\Sigma$-equations and those generated from the semilattice equations on $\Sigma^+$-terms do exist.  
Let $\mathcal{S}$ be the set equations of Figure \ref{fig:string-equations},  \textit{i.e.},  the semilattice equations over $\Sigma^+$-terms. 
%Let $\mathcal E$ be an arbitrary set of $\Sigma$-equations. 
Then we have the following result. 
\begin{proposition}[Soundness]
\label{prop:soundness}
The category $\WellTypedMdaEcospans/{\mathcal{S}}$ is a semilattice-enriched PROP. 
\end{proposition}
Note that the symmetric monoidal equations are absorbed in the hypergraph representation,  but the semilattice equations are covered by the fact we have quotiented our representation by certain rewrites.

The reason for calling the previous result \textit{soundness} is that the interpretation function on $\Sigma^+$-terms induces a semilattice-enriched SMC functor $\llbracket - \rrbracket: \textbf{PROP}^+(\Sigma) \to \WellTypedMdaEcospans/{\mathcal{S}}$,  which further induces the following semilattice-enriched SMC functor,  for a set of equations $\mathcal E$:
\[
	\llbracket - \rrbracket_{\mathcal E}: \textbf{SMT}^+(\Sigma, \mathcal E) \to \WellTypedMdaEcospans/{\mathcal{S,E}}
\]
which by uniqueness must be the free semilattice-enriched functor from $\textbf{SMT}^+(\Sigma, \mathcal E)$ induced by the interpretation of $\Sigma$ as itself.  Unfolding definitions,  it follows that
\[
	f = g \quad \Rightarrow \quad \llbracket f \rrbracket \Rrightarrow^{*}_{\mathcal{S}, \mathcal{E}} \llbracket g \rrbracket~ . 
\]
Note that it is immediate from the previous result that $\WellTypedMdaEcospans/{\mathcal{S,E}}$ is indeed a semilattice-enriched PROP.

The equations of Figure~\ref{fig:string-equations} -- excepting the commutativity equation -- can all be directed in an obvious way to form a terminating rewrite system,  which results in a cospan in \textit{weak normal form}. 
\begin{proposition}[Weak normal form]
\label{prop:wnormal_form}
For each cospan $f$ in ${\WellTypedMdaEcospans}/{\mathcal{S}}$ there is a \textit{weak normal form} such that 
\[
	f = f_1 + \ldots + f_n
\]
 such that each $f_i$ contains no hierarchical edges,  and for all $i \neq j$ we have $f_i \neq f_j$.
\end{proposition}    
The proof follows the familar argument (\textit{i.e.},  termination of distributivity rewrites) that $\Sigma^+$-terms can be rewritten via the semilattice-enrichment equations into a similar normal form.

This normal form can be used to prove the main theorem of the paper: the (full) completeness of our combinatorial representation of $\textbf{SMT}^{+}(\Sigma, \mathcal {E} )$. 
\begin{theorem}[Full completeness]
We have the following equivalence of categories
\[
	\textbf{SMT}^{+}(\Sigma, \mathcal {E} ) \cong \WellTypedMdaEcospans / \mathcal{S,E}~.
\]
\end{theorem}
\begin{proof}
The equivalence is given by the functor $\llbracket - \rrbracket$.  First, note that the functor is surjective on objects,  so it suffices to prove the functor is both full and faithful.  In each case,  the result follows by a normal form argument in combination with relying on the equivalence $(*)$ $\textbf{SMT}(\Sigma, \mathcal E) \cong \textbf{MHypI}(\Sigma)/\mathcal{E}$ of symmetric monoidal theories and (appropriately quotiented) plain MDA cospans which is given by the restriction of $\llbracket - \rrbracket$ to $\textbf{SMT}(\Sigma, \mathcal E)$ as a subcategory of $\textbf{SMT}^{+}(\Sigma, \mathcal {E} )$ (note that cospans embed in extended cospans when the internal interface is isomorphic to the external interface); see Theorem 35 of ~\cite{bonchi_string_2022-2}.

@@

Unfolding the definitions,  we can spell out the previous theorem in terms of EDPOI-rewriting. 
\begin{corollary}[Completeness of rewriting]
For any $f, g \in \textbf{SMT}^{+}(\Sigma, \mathcal{E})$,  we have
\[
	f = g \quad \iff \quad \llbracket f \rrbracket \Rrightarrow^{*}_{\mathcal{S}, \mathcal{E}} \llbracket g \rrbracket~ . 
\]
 \end{corollary}
%
%\begin{theorem}[Theorem 35~\cite{bonchi_string_2022-2}]
%
%    $a \Rightarrow_{R} b$ iff $\llangle {}^{\ulcorner} a {}^{\urcorner} \rrangle \Rrightarrow_{\llangle {}^\ulcorner R {}^\urcorner \rrangle} \llangle {}^{\ulcorner}b {}^{\urcorner} \rrangle$.
%    Where $\llangle \cdot \rrangle : \textsf{PROP}(\Sigma) + \textsf{Frob} \to \catname{Csp_{D}(Hyp_{\Sigma})}$ and $\Rrightarrow$ is a convex rewriting relation for morphisms in $\catname{Hyp_{\Sigma}}$.
%\end{theorem}
%
%\begin{proposition}
%    $\llangle {}^{\ulcorner} a {}^{\urcorner} \rrangle \Rrightarrow_{\llangle {}^\ulcorner R {}^\urcorner \rrangle} \llangle {}^{\ulcorner}b {}^{\urcorner} \rrangle$ iff $[\mathcal{J}(a)] \Rrightarrow_{[R]}^{+} [\mathcal{J}(b)]$.
%     Note that $a$ and $b$ are from $\textsf{PROP}(\Sigma)$.
%\end{proposition}
%
%\begin{theorem}
%    $a \Rightarrow_{\langle l, r \rangle}^{+} b$ in $\textsf{PROP}^{+}$ iff $[a] \Rrightarrow_{\langle [l],[r] \rangle} [b]$ in $\sfrac{\WellTypedMdaEcospans}{\mathcal{S}}$.
%\end{theorem}
Note that in string diagram rewrite theory as developed in \cite{bonchi_string_2022-2},  some extra steps are taken to prove the claim that $f \rightsquigarrow g$ if and only if $\llbracket f \rrbracket \Rrightarrow^{*}_{\mathcal{S}, \mathcal{E}} \llbracket g \rrbracket$,  for an appropriate definition of $\Sigma$-term rewriting $\rightsquigarrow$.  We omit this extra detail for brevity and clarity,  dealing with equational theories rather than rewrite theories,  but it is straightforward to extend our results to prove an analogous claim.
